The importance of alternatives in scalar implicature

### 2016-01-21 Paper/Paper.tex
title tweaks
updated draft. reordered/rewritten intro and modeling section

\title{The importance of negative alternatives in pragmatic implicature}

\author{{\large \bf Benjamin Peloquin} \\ \texttt{bpeloquin@stanford.edu} \\ Department of Psychology \\ Stanford University \And {\large \bf Michael C. Frank} \\ \texttt{mcfrank@stanford.edu} \\ Department of Psychology \\ Stanford University}

@@

\begin{abstract}
Frameworks of pragmatic enrichment, starting with Grice 1975, implicitly
assume the presence of salient alternatives. For example, the standard
quantity implicature arises the presence of a competiting and more
informative alternatives. Using a Bayesian formulzation of this theory
we assess the degree to which alternatives influence pragmatic
enrichment of scalar items. Our findings suggest that the presence of a
complete of set of possible alternatives.

@@

Implicature as a case study for pragmatics

RSA as a way of studying implicature ({\textbf{???}})

Scalar implicature in particular, what are the alternatives?
({\textbf{???}}) ({\textbf{???}})

Using the scalar diversity approach to get beyond the some/all
implicature ({\textbf{???}})

Outline of the rest of the paper

@@

\subsection{Participants}\label{participants}

@@

\subsection{Participants}\label{participants-1}

@@

\section{Experiment 3a,b: Adding Negative
alternatives}\label{experiment-3ab-adding-negative-alternatives}

\subsection{Procedure}\label{procedure-2}

\section{Model}\label{model}

\subsection{Details}\label{details}

\subsection{Model fitting and
comparison}\label{model-fitting-and-comparison}

@@

@@

Thanks to NSF BCS \#XYZ. Thanks to Michael Franke, Judith Degen, and
Noah Goodman.

@@

>>>>

\title{The importance of alternatives in scalar implicature}

\author{{\large \bf Benjamin Peloquin} \\ \texttt{bpeloqui@stanford.edu} \\ Department of Psychology \\ Stanford University \And {\large \bf Michael C. Frank} \\ \texttt{mcfrank@stanford.edu} \\ Department of Psychology \\ Stanford University}

@@

\begin{abstract}
Succesful communication regularly requires listeners to make pragmatic
inferences - enrichments beyond the literal meaning of a speaker's
utterance. For example, listeners routinely enrich the meaning of
``some'' to ``some, but not all'' when interpreting a sentence such as
``Bob ate some of the cookies.'' A Gricean account of this phenomena
assumes the presence of \emph{salient alternatives} with varying degrees
of informativity. ``Some,'' in the example above, is enriched to ``some,
but not all'' in the presence of the stronger alternative ``all.''
Empirical evidence for the presence of such alternatives and accounts of
their effect on implicature has been limited. Our current work explores
the role different scale representations may play in scalar implicature
using emperical measures of literal semantics and a Bayesian model of
implicature generation. Comparisons with human judgments indicate that
pragmatic inference may rely on fairly complete alternative sets rather
than the typical entialment only scales assumed in most research to
date.

@@

Humpty Dumpty Quote\ldots{}

Successful communication regularly requires listeners to make inferences
that go beyond the literal semantic content of a speaker's utterance.
``Scalar implicature'' is a hallmark of such pragmatic enrichment. For
example, listeners routinely enrich the meaning of the scalar item
``some'' to ``some, but not all'' in sentences like ``Bob ate some of
the cookies'' (van Tiel, 2014). A Gricean account of this phenomena
assumes listeners reason about the intended meaning of a speaker while
incorporating knowledge about a) alternative scalar items a speaker
could have used (such as ``all'') and b) the relative informativity of
using such alternatives (Grice, 1975). Under the Gricean account, a
listener will infer that the speaker must have intended that Bob did not
eat ``all'' the cookies because it would have been
\emph{underinformative} for the speaker to use ``some'' when ``all''
could have been used.

The presence of ``salient" alternatives is critical to the standard
Gricean account of scalar implicature (Grice, 1975). Recent experimental
evidence appears to support this. Using a gumball paradigm, Degen \&
Tanenhaus (2014) demonstrated that the scalar item \emph{some} was
judged less appropriate (natural) for small sets of items when exact
numbers were seen as viable alternatives. This finding points to the
impact a particular alternative set can have on judgments regarding
scalar appropriateness. Using a fundamentally differnt approach, Franke
(2014) formalized the Gricean account of implicature using a Bayesian
model, modeling graded typicality judgments and scalar implicature.
Results supported `none', `one' and `all' as the most serious (salient)
alternatives to ``some.''

Taken together, each of the previous studies signal the importance of
scale representation (set of available alternatives) in scalar
implicature. However, like many recent investigations which have focused
only on a small subset of scalar items, both Degen \& Tanehaus (2014)
and Franke (2014) focus exclusively on the scalar family ``some/all.''
This lack ``Scalar Diversity'' (van Tiel, 2014) makes generalization
across different scalar families problematic. van Tiel (2014), found
substantial variation in the rate of upper-bounded construals among over
40 different scalar pairs. This finding suggests that implicature may
differ widely between different scales. We adopt a ``Scalar Diversity''
approach in our experimental design in order to extend our investigation
beyond ``some/all'' scalar items. In the following set of studies we
examine implicature for ``some/all'' and five additional scalar families
from a range of grammatical classes (see Figure 5 for a complete set of
scalar items used). The set of scalar items for this study were chosen
from among the 40 used in van Tiel (2014).

Consider a scalar item other than ``some'' - in the context of a scalar
item such as ``good'', the idea of set of lexical alternatives certainly
xfeels intuitive (``bad'', ``excellent'', etc.). But measuring the
extent to which these alternatives are salient is not so clear. Nor is
how to measure their impact on pragmatic intepretations of the scalar
item ``good.'' In an experimental setting, querying a participant about
the saliency of a particular alternative is problematic because the
alternative must be made salient during the query. In the spirit of
Franke (2014) and Degen \& Tanenhaus (2014), we pair a computational
model of implicature with empirical measurments of linguistic material.
This combination allows us to simulate the effect of various scale
representations on implicature. In particular, we adopt a Bayesian model
of implicature and a food-review paradigm to empirically quantify
literal semantics and pragmatic judgments for our scalar stimuli.

Bayesian models of pragmatic enrichment have accurately predicted human
judgments in ad-hoc (Frank \& Goodman, 2012) and embedded (Stulhumuller
\& Goodman, 2014) implicature settings. Our current model follows both
these studies and is based on Rational Speech-act theory (RSA). RSA
frames language understanding as a special case of social cognition
(Stulhumuller \& Goodman, 2014), in which listeners and speakers reason
about one-another. In this framework, a listener uses Bayesiain
inference to interpret the intended meaning \(m\) of a speaker who has
made an utterance \(u\). We will make the model specifications explicit
later in the paper.

In the following study we investigate the role of scale representation
on scalar implicature. To do so we combine a computational framework
with empirical measurements to simulate the type of scale
representations that might be available to a listener. In the following
sections we outline the specifics of the model and the particular
components for which we need to to make empirical measurements. We then
outline a series of experiments meant to a) quantify literal semantics
for otherwise ambiguous scalar items, b) measure pragmatic judgments and
c) generate a set of plausible alternatives for our set of scalar pairs
taken from van Tiel (2014). In particular, we adopt a food review
paradigm throughout our empirical data gathering, quantifying linguistic
meaning (both literal semantics and pragmatic judgments) as
distributions over star ratings. Finally, we simulate the impact of
three different scale representations, ``entailment'', ``mid'' and
``symmetric'' on implicature generation, comparing model predictions
with pragmatic judgments captured empirically. Results indicate that
humans may employ more ``fully specified'' scale representatinos,
relying on the presence of multiple alternatives during pragmatic
enrichment.

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

\section{Modeling implicature: Rational Speech-act
theory}\label{modeling-implicature-rational-speech-act-theory}

We adopt a Bayesian model of scalar implicature known as Rational
Speech-act theory (RSA). In particular, RSA frames language
understanding as a special case of social cognition (Frank \& Goodman,
2012; Stuhmuller \& Goodman, 2014) in which listener and speaker agents
reason about one another. We focus on the problem of a listener
inferring the meaning of a speaker's utterance. Let \(u\) be an observed
utterance made by our speaker with an intended meaning \(m\). We assume
that the listener has access to a space of possible meanings \(M\) and
some model relating meanings \(m \in M\) to \(u\).

Upon hearing an utterance \(u\) the Listener agent evaluates all
candidate word meanings, computing their posterior probabilities
\(p_{L1}(m | u)\). This quantity is proportional to the product of the
prior probability \(p(m)\) of a particular meaning and the likelihood
\(p(u | m)\).

\[p_{L1}(m | u) = \frac{p(u | m)p(m)}{p(u)}\]
\[= \frac{p(u | m)p(m)}{\sum_{m \in M}p(u | m)}\]

The \textbf{prior} \(\mathbf{p(m)}\) represents the listener's
expectations about plausible meanings (star ratings), \emph{independent}
of the utterance \(u\). To avoid biasing the data we assume priors over
meanings are uniform, however future investigations may want to
experiment with different priors, either by inferring them or using
empirical measures

\(\mathbf{p(u | m)}\) quantifies the \textbf{likelihood} that a speaker
would make a particular utterance \(u\) given an intended meaning \(m\).
Our model assumes that speakers choose words to maximize the utility of
an utterance in context. Utility is operationalized as the informativity
of a particular utterance (suprisal) minus a cost. In particular, the
speaker chooses an utterance in order to maximize the utility of a
listener who interprets her utterance literally \(p_{L0}(m|u)\). \(L_0\)
denotes a literal listener interpretation.
\[p(u | m) = \frac{e^{-\alpha(-log(p_{L0}(m|u)) - cost(u))}}{\sum_{m \in M}e^{-\alpha(-log(p_{L0}(m|u)) - cost(u))}}\]

The \textbf{posterior} \(\mathbf{p_{L1}(m | u)}\) quantifies a
listener's pragmatic enrichment of an utterance \(u\) as degree of
belief that \(m\) is the intended meaning of the speaker, given an
utterance \(u\). \(L_1\) denotes a pragmatic listener interpration.

\subsection{Populating the model: Measuring literal semantics and
pragmatic
judgments}\label{populating-the-model-measuring-literal-semantics-and-pragmatic-judgments}

In order to simulate the role of scale representation on implicature we
pair a computational framework with empirical data. In particular we use
experimental tools to populate three components of the model. First, to
measure otherwise ambigous literal semantics \(\mathbf{p_{L0}(m|u)}\)
(our ``Scalar Diversity''" approach). Second, to generate a set of
plausible alternatives to the original scalar pairs taken from van Tiel
(2014). Lastly, to obtain human pragmatic judgments for model comparison
\(\mathbf{p_{L1}(m|u)}\).

Starting with the first component, consider the scalar items ``some''
and ``all.'' Both are convenient for modeling in a computational
framework like the one we've proposed above. Their literal semantics
\(\mathbf{p_{L0}}\) are easily quantified over sets.

\[\text{\textit{some of the A's are B's.}}\]
\[[[some]] = \{<A, B> : |A \cap B| \geq 1 \}\]

\[\text{\textit{all of the A's are B's.}}\]
\[[[all]] = \{<A, B> : \frac{|A \cap B|}{|A|} == 1 \}\]

The same cannot be said for items such as ``good / excellent'',
``memorable / unforgettable'' or ``liked / loved.''

In the following set of studies we adopt a food-review paradigm to
measure literal semantics and pragmatic judgments empirically. In order
to capture graded typicality judgments as in Degen \& Tanenhaus (2014)
we use a compatiblity measure dependent variable, pairing scalar items
with star-ratings in our literal listener task (Experiments 1a, 3a and
4). Data from these experiments is used to approximate the literal
listener semantics for a given scalar item \(u\) as a distribution over
star-ratings \(m \in \text{\{1 - 5 stars\}}\). Similarly, pragmatic
judgment data from Experiments 1b and 3b serve as our primary points of
comparison with model predictions \(\mathbf{p_{L1}(m|u)}\), again
quantifying judgments as distributions over star-ratings. In the
following section we take the read through our experiments.

\begin{CodeChunk}
\begin{figure*}[t]

{\centering \includegraphics{figs/stimuli_exp1-1}

}

\caption[The left panel shows a trial from Experiment 1a with the target scalar `liked']{The left panel shows a trial from Experiment 1a with the target scalar `liked'. Participants responses to the binary dependent variable are used to quantify literal semantics. The right panel shows a trial from Experiment 1b with the target scalar `liked'. Participants were asked to generate the star rating they think the speaker likely intended, given the target scalar (in this case 'liked').}\label{fig:stimuli_exp1}
\end{figure*}
\end{CodeChunk}

@@

Experiment 1a and 1b were conducted to approximate literal listener
semantic distributions (1a) and pragmatic judgments (1b) for five pairs
of scalar items taken from van Tiel (2014) (see figure 5 for stimuli).
Each pair consists of a ``weaker'' scalar term (e.g. ``some'') and a
stronger scalar term (e.g. ``all''). We are calling these ``entialment
scales''``, because they encode the assumption that a listener need only
have a the stronger alternative present to generate an enriched
interpretation of the weaker term.

\subsection{Methods}\label{methods}

\subsubsection{Participants}\label{participants}

Participants for both tasks were recruited on Amazon Mechanical Turk and
paid \$0.20 for their participation. Thirty participants were recruited
for Experiment 1a. Data for N participants was thrown out after
participants either failed to pass two training trials or were not
native English speakers, leaving a total sample of N. Fifty participants
were recruited for experiment 1b. Data for N participants was thrown out
after participants either failed to pass two training trials or were not
native English speakers, leaving a total sample of N.

\subsubsection{Design and procedure}\label{design-and-procedure}

The left panel of Figure 1 shows a screen shot of a trial from
Experiment 1a. Participants were presented with a target scalar item and
a star-rating (between 1-5 stars) and asked to judge the compatibility
of the scalar item and star-rating. Compatibility was assessed through a
binary ``yes/no'' response to a question of the form ``Do you think that
the person \_\_\_ed the food?" where a target scalar was presented in
the ``\_\_\_ed." Each participant was presented all scalar item and
star-rating combinations with randomization.

The right panel of Figure 1 shows a screen shot of a trial from
Experiment 1b. On each trial, participants were presented with a
one-sentence prompt containing a target scalar item such as ``Someone
said they \_\_\_ed the food." They were then asked to generate a
star-rating according to what they think the reviewer likely gave. Each
participant was presented all scalar items with randomization.

The left panel of figure two shows literal semantics for each of the
scalar terms ``liked'' and ``loved'' as distributions over star-ratings.

Chi-squared test? Null results, not really important here anyway\ldots{}

\begin{CodeChunk}
\begin{figure*}[t]

{\centering \includegraphics{figs/exp1Plots-1}

}

\caption[The left panel plots literal semantic distributions for the scalar pair 'liked/loved']{The left panel plots literal semantic distributions for the scalar pair 'liked/loved'. The y-axis is the percentage of 'yes' responses for a scalar term that is compatibible with the number of stars on the x-axis (ie 100\% of respondents beleived that 'liked' was compatible with both 4 and 5 stars). Erro bars are 95\% confidence intervals. The right panel plots pragmatic judgments for the scalar pair 'liked/loved'. The y-axis is the proportion of selection of the star-rating on the x-axis (ie over 80\% of judgments when prompted with 'loved' were 5-stars).}\label{fig:exp1Plots}
\end{figure*}
\end{CodeChunk}

In Experiment 2 we asked participants to generate plausible alternatives
for the scalar items presented in Experiment 1.

\subsection{Methods}\label{methods-1}

\subsubsection{Participants}\label{participants-1}

Using Amazon's Mechanical Turk, N workers were paid \$0.20 to
participate. n workers were dropped from the analysis because they
either failed to pass training trials or were not native English
speakers. The final sample was N workers, all of whom were naive to the
purpose of the experiment.

\subsubsection{Design and procedure}\label{design-and-procedure-1}

Participants were presented a target scalar term embedded in a sentence
such as, ``In a recent restaurant review someone said they thought they
\emph{----}ed the food.'' Participants were then asked to generate
plausible alternatives by responding to the question, ``If they'd felt
differently about the food, what other words could they have used
instead of \emph{----}?'' and asked to generate three alternatives.

Figure 3 plots the combined counts of alternatives generated for the
scalar items ``liked'' and ``loved.'' Alternative distributions for the
other scalar pairs (e.g.. ``good/excellent'',
``memorable/unforgettable'') were similarly long-tailed. In Experiments
3a,b we took the top two alternatives generated for each scalar item to
enrich the entailment only scales from Experiments 1a,b. In experiment 4
we further enriched the scales from Experiments 3a,b with a neutral
valence scalar chosen from the alternative sets.

\begin{CodeChunk}
\begin{figure*}[t]

{\centering \includegraphics{figs/exp2_altsPlot_likedLoved-1}

}

\caption[Caption goes here]{Caption goes here}\label{fig:exp2_altsPlot_likedLoved}
\end{figure*}
\end{CodeChunk}

\section{Experiment 3a,b: Incorporating top
alternatives}\label{experiment-3ab-incorporating-top-alternatives}

Experiment 3a,b are identical to Experiments 1a,b except the set of
target scalar was expanded to include the top two alternatives generated
for each scalar family. The orange colored items in table 1 denote
additional scalar items added in Experiments 3a,b.

Participants for both studies were recruited on Amazon Mechanical Turk
and paid \$0.20 for their participation. Thirty participants were
recruited for Experiment 3a (Literal Listener task). Data for N
participants was thrown out after participants either failed to pass two
training trials or were not native English speakers, leaving a total
sample of N.

Fifty participants were recruited for Experiment 3b (Pragmatic Listener
task). Data for N participants was thrown out after participants either
failed to pass two training trials or were not native English speakers,
leaving a total sample of N.

@@

The procedure of Experiments 3a,b follow the same form as Experiments
1a,b with the expanded set of target scalar items.

Distributional differences.

\section{Experiment 4: Symmetric
scales}\label{experiment-4-symmetric-scales}

Experiment 4a is identical to Experiments 1a and 3a except the set of
target scalar was expanded to include a neutral valence scalar item for
each scalar family. The purple colored items in Table 1 denote
additional scalar items added in Experiments 4.

\subsection{Participants}\label{participants-3}

Participants for both studies were recruited on Amazon Mechanical Turk
and paid \$0.20 for their participation. Thirty participants were
recruited for Experiment 4 (Literal Listener task). Data for N
participants was thrown out after participants either failed to pass two
training trials or were not native English speakers, leaving a total
sample of N.

@@

The procedure of Experiment 4 follow the same form as Experiments 1a and
3a with the addition of a neutrally valence scalar item.

\subsection{Results and Discussion}\label{results-and-discussion-3}

Distributional differences.

\section{Temp section Model was here}\label{temp-section-model-was-here}

\subsubsection{Results}\label{results}

Table here

Figure 4 plots model / human correlation for our three simulations.
Alpha was tuned separately for each of these runs\ldots{}

\begin{CodeChunk}
\begin{figure*}[t]

{\centering \includegraphics{figs/performancePlots-1}

}

\caption[The left panel shows improved model fit as scale representations are enriched with more scalar items]{The left panel shows improved model fit as scale representations are enriched with more scalar items. Correlations are coputed using pragmatic judgment data from Experiments 1b and 3b. The right panel plots model predictions using full symmetric scales versus human judgments from Experiment 3b.}\label{fig:performancePlots}
\end{figure*}
\end{CodeChunk}

@@

By varying the type of scale representations available to our Bayesian
model we investigated the effects of alternatives on scalar implicature.
Model fit with human judgement was significantly improved by the
inclusion of alternatives beyond the typical ``strong'' and ``weak''
scalar items. In fact, we found that both neutral and negative valence
scalar items contribute to human-like implicature generation within our
framework.

\textbackslash{}begin\{CodeChunk\}
\textbackslash{}begin\{figure\}{[}h{]}

\{\centering \includegraphics{figs/allScalesTable-1}

\}

\textbackslash{}caption{[}This table shows the stimuli used in
Experiments 1a,b, 3a,b and 4{]}\{This table shows the stimuli used in
Experiments 1a,b, 3a,b and 4. Colors denote the additions made in each
experiment. For example, in Experiment 1a we measured literal semantics
for `liked/loved'. In Experiment 3a,b we extended this group to
`textit\{hated/disliked/liked/loved'. In Experiment 4 we extended this
grou to `hated/disliked/felt indifferent
about/liked/loved'.\}\label{fig:allScalesTable}
\textbackslash{}end\{figure\} \textbackslash{}end\{CodeChunk\}

@@

Thanks to NSF BCS XYZ. Thanks to Michael Franke, Judith Degen, and Noah
Goodman.

@@

use a food review paradigm in which `meaning' for a given scalar item
was quantified as a distribution over star-ratings. Quantifying both
literal semantic content and pragmatic judgments as distributions over
stars provides an important interface to our model, especially for
scalar items with ambigous literal semantics (e.g.~items such as
``liked/loved'' or ``memorable/unforgettable''). In addition to
quantifying literal semantics and pragmatic judgments in Experiments
1a,b and Experiments 3a,b, Experiment 2 used a modified cloze task,
based on van Tiel (2013, Experiment 2???), to generate a set of
plausible alternatives to the scalar items used in Experiments 1a,b. We
extend the set of alternatives measured in Experiments 3a,b and 4 using
data from Experimeng 2. Additionally, by looking at the relative
frequencies of alternatives generated in Experiment 2 we were able to
compute a saliency measure for each alternative. Using the literal
semantic data from Experiments 1a, 3a and 4, we modeled pragmatic
enrichment using RSA and compared model predictions to human judgments
obtained in Experiments 1b and 3b. Model performance was significantly
improved with the incorporation of alternatives. This result suggests
that more ``fully specified" scale representations may be active for
human participants while making pragmatic judgments for scalar items.
